---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MedBounds

<!-- badges: start -->
<!-- badges: end -->

The MedBounds package implements the methodology from the paper ["Testing
Mechanisms"](https://www.jonathandroth.com/assets/files/TestingMechanisms_Draft.pdf)
by Soonwoo Kwon and Jonathan Roth. The package provides tests for the "sharp
null of full mediation", which conjectures that the effect of a treatment
operates through a particular conjectured mechanism (or set of mechanisms) M. It
also provides lower bounds on the fraction of "always-takers" who are affected
by the treatment despite having the same value of M regardless of treatment
status.

## Installation

You can install the development version of MedBounds from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# Install devtools if not already installed
install.packages("devtools")

# Install package
devtools::install_github("jonathandroth/MedBounds")
```

## Application to Baranov et al. (2020)

We illustrate how the package can be used by walking through how the code can be
applied to the application of Baranov et al. (2020) in Section 5.2 of ["Testing
Mechanisms"](https://www.jonathandroth.com/assets/files/TestingMechanisms_Draft.pdf). In
Baranov et al. (2020), $D$ is a treatment for depression and $Y$ is an index of
outcomes for women's financial empowerment. We are interested in whether the
effect of $D$ on $Y$ can be explained by a mediator, or set of mediators,
$M$. We consider three choices for $M$: (a) the presence of a grandmother in the
home, (b)relationship quality with the woman's husband, and (c) the combination
of these two mechanisms.


We start with loading the required packages and the data.
```{r, message = F, warning=F}
# Load MedBounds
library(MedBounds)

# Load other packages that are required to run the example
library(dplyr)
library(ggplot2)
library(haven)

# Load data
data("baranov_data")

# Restrict to the experimental sample
mother_data <- mother_data %>% filter(THP_sample == 1)
```

We begin with the case where $M$ is a binary indicator for the presence of a
grandmother in the home. The outcome variable $Y$ is continuous, and for ease of
transparency and for conducting inference, we discretize the outcome into 5 bins
based on the unconditional quantiles of the outcome. As noted in the paper, the
test still remains valid under such a discretization but potentially loses
sharpness. With some abuse of notation, we from now on write $Y$ to refer to
this discretized outcome. 

The main functions we will be using are: 1) `partial_density_plot()` to plot the
partial densities to visually detect potential violations of the *sharp null*,
2) `test_sharp_null()` to conduct a statistical test for the *sharp null* and 3)
`lb_frac_affected()` to compute the sharp lower bound for the fraction of
always-takers (or never-takers) that are affected by treatment. While this
example covers the basic usage of these function, please refer to the
documentation of each function (e.g., `?test_sharp_null`) for a more detailed
description. 

### Graphical Evidence

We first provide graphical evidence using a partial density plot. Since such
(partial) density plots are often helpful to understand where the violations of
the *sharp null* is coming from, the package includes a function
`partial_density_plot()` that can be used to produce such plots. The following
snippet reproduces Figure 3 of the main paper.

```{r}
nt_plot <-
  partial_density_plot(df = mother_data,
                       d = "treat",
                       m = "grandmother",
                       y = "motherfinancial",
                       num_Ybins = 5,
                       plot_nts = T,
                       density_1_label = "Prob. In Treated Group (P(Y,M=0|D=1))",
                       density_0_label = "Prob. In Control Group (P(Y,M=0|D=0))") +
  ylab("Probability") +
  xlab("Event: No Grandmother Present (M=0) and Y in Stated Range")

nt_plot +
  annotate(geom = "text", 
           x = 5,y=0.15,
           label = "Higher prob. in \n treated group, \n violating sharp null", size = 3) +
  geom_segment(x=4.7,y=0.17,
               xend = 4.5, yend = 0.175,
               arrow = arrow(length = unit(0.03,"npc")),
               color = "black", show.legend = F) +
  
  geom_segment(x=5,y=0.13,
               xend = 5, yend = 0.10,
               arrow = arrow(length = unit(0.03,"npc")),
               color = "black", show.legend = F) +  
  theme(legend.position="bottom",
        legend.title = element_blank())
```

This figure shows estimates of $P(Y=y,M=0 \mid D=d)$ for both $d=1$ and
$d=0$. Under monotonicity, then as shown in Section 2 of the paper, we should
have that $P(Y=y, M=0 \mid D=1) \leq P(Y=y, M=0 \mid D=0)$ for all values of
$y$. As shown in the figure, however, this inequality appears to be violated at
large values of $y$, suggesting that the outcome for some treated never-takers
improved when receiving the treatment.

### Testing the *sharp null of full mediation*

While the figure above hints a possible violation of the *sharp null*, it does
not come with any uncertainty quantification. The function `test_sharp_null()`
conducts statistical inference of the *sharp null* using the method described in
Section 4 of the paper. The following snippet runs this test based on Cox and
Shi (2023) (as can be seen from `method = "CS"`).

```{r, cache = TRUE}
test_result <- test_sharp_null(df = mother_data,
                               d = "treat",
                               m = "grandmother",
                               y = "motherfinancial",
                               method = "CS",
                               num_Ybins = 5,
                               cluster = "uc",
                               analytic_variance = TRUE,
                               refinement = TRUE)

test_result$pval
```

The test gives a p-value of `r round(test_result$pval, digits = 3)`, and thus
the *sharp null* is rejected at a 5% significance level. Here, the p-value
corresponds to the smallest value of $\alpha$ for which the test rejects. 


### Calculating the lower bound on fraction of always-takers affected by outcome

We now compute lower bounds on the fraction of always-takers whose outcome is
affected by the treatment despite having the same value of $M$ under both
treatments. This gives a sense of the strength of mechanisms other than $M$:
they tell us what fraction of the $k$-always takers have a direct effect of the
treatment. The definition of the lower bounds can be found in Section 3.1 of the
paper. The function `lb_frac_affected` computes (a point estimate of) this lower
bound. The argument `at_group = 0` corresponds to computing this lower bound for
$0$-always takers, or, the never-takers. 
```{r, cache = TRUE}
lb_nts <- lb_frac_affected(df = mother_data,
                           d = "treat",
                           m = "grandmother",
                           y = "motherfinancial",
                           num_Ybins = 5,
                           at_group = 0)
lb_nts
```
Our estimates of the lower bound imply that at least `r round(100 * lb_nts)`
percent of never-takers are affected by the treatment.

### Allowing for defiers

By default, MedBounds imposes the monotonicity assumption that the treatment can
only increase the value of $M$. In this setting, this means that everyone who
would have a grandmother present without receiving CBT treatment would also have
one present when receiving CBT treatment. We can relax this assumption by
setting the `max_defiers_share` parameter to be non-zero, which bounds the
number of "defiers" by `max_defiers_share`. 

We rerun the test above with
`max_defiers_share = .01`, which allows one percent of the population to be defiers.
```{r, cache = TRUE}
test_result_defiers <- test_sharp_null(df = mother_data,
                                       d = "treat",
                                       m = "grandmother",
                                       y = "motherfinancial",
                                       method = "CS",
                                       num_Ybins = 5,
                                       cluster = "uc",
                                       analytic_variance = TRUE,
                                       max_defiers_share = .01,
                                       refinement = TRUE)

test_result_defiers$pval
```
The p-value increases to `r round(test_result_defiers$pval, 3)`, so the test
rejects the *sharp null* even if you allow one percent of the population to be
defiers. 

Likewise, we can also calculate the lower bound on the fraction of never-takers
under this relaxed monotonicity.
```{r, cache = TRUE}
lb_nts_defiers <- lb_frac_affected(df = mother_data,
                                   d = "treat",
                                   m = "grandmother",
                                   y = "motherfinancial",
                                   num_Ybins = 5,
                                   at_group = 0,
                                   max_defiers_share = .01)
lb_nts_defiers
```
Our estimates of the lower bound imply that at least `r round(100 * lb_nts_defiers)`
percent of never-takers are affected by the treatment when we allow one percent of the
population to be defiers.


### Results for an alternative mechanism (relationship quality with husband)

We next turn to the setting where we are interested in testing whether the
effect is mediated by relationship quality with the husband, which is measured
on a 1-5 scale. We can again test the *sharp null* and estimate a lower bound on
the fraction affected.

```{r, cache = TRUE}
test_sharp_null(df = mother_data,
                d = "treat",
                m = "relationship_husb",
                y = "motherfinancial",
                num_Ybins = 5,
                method = "CS",
                cluster = "uc",
                analytic_variance = TRUE)$pval
```
Again, we reject the *sharp null* that all the treatment effect goes through the
relationship quality with the husband.

We also estimate a lower bound on the fraction of always-takers:
```{r, warning = FALSE}
lb_frac_affected(df = mother_data,
                 d = "treat",
                 m = "relationship_husb",
                 y = "motherfinancial",
                 num_Ybins = 5,
                 at_group = NULL,
                 allow_min_defiers = TRUE)
```
Here, the parameter `at_group = NULL` makes the function to compute the lower bound
on the fraction of always-takers, pooled across
different $M$ values to be around 10 percent. The parameter `allow_min_defiers = TRUE`
is used to allow a minimal number of defiers for monotonicity to be consistent
with the data -- see footnote 25 of the paper for details.

### Combination of both mechanisms

Finally, we test the null hypothesis that the treatment effect is explained by
the combination of the two mechanisms.
```{r, cache = TRUE}
test_result_both <- test_sharp_null(df = mother_data,
                                    d = "treat",
                                    m = c("relationship_husb",
                                          "grandmother"),
                                    y = "motherfinancial",
                                    num_Ybins = 5,
                                    method = "CS",
                                    cluster = "uc",
                                    analytic_variance = TRUE)

test_result_both$pval
```
With a p-value of `r round(test_result_both$pval, 3)`, 
we cannot reject the sharp null that the combination of presence of grandmother and
relationship quality with husband fully explain the treatment effect. 
	
Again, we can estimate a lower bound on the fraction of those affected by
treatment, pooled across different $M$ values:
```{r, warning = FALSE}
lb_frac_both <- lb_frac_affected(df = mother_data,
                                 d = "treat",
                                 m = c("relationship_husb",
                                       "grandmother"),
                                 y = "motherfinancial",
                                 num_Ybins = 5,
                                 allow_min_defiers = TRUE)
lb_frac_both
```
We estimate a lower bound of `r round(100 * lb_frac_both)` percent, but this does not appear
to be statistically significant given the test result above.




